---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading of packages
```{r message=F, warning=F}
library(lubridate)
library(plyr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(lattice)
library(knitr)
```
## Loading and preprocessing the data
### Step 1: Loading and preprocessing the data
### Loading and preprocessing the data

#### 1.1. Load the data (i.e. read.csv() )
###### check the data
```{r}
readLines("activity.csv", 4)
```
###### read the data as csv
```{r}
data <- read.csv("activity.csv", header = TRUE)
head(data)
```
###### check the class of the data
```{r}
class_data <- lapply(data, class)
class_data
```


## What is mean total number of steps taken per day?
### Step 2: Calculation of the mean total number of steps taken per day

#### 2.1 Calculation of sum of the steps per day ignoring the NA values
#### The value assigned to the NA elements (missing values) is zero in this implementation
#### This is accomplished through the na.rm = TRUE argument

```{r}
total_steps_per_day_data <- data %>% 
        group_by(date) %>% 
        summarise(steps_per_day = sum(steps, na.rm = TRUE)) %>%
        ungroup


head(total_steps_per_day_data)
str(total_steps_per_day_data)
head(total_steps_per_day_data$steps_per_day)
```

#### 2.2 Creation of a histogram of the number of steps

```{r}
hist(total_steps_per_day_data$steps_per_day, breaks = 8, xlab = "Total Steps per Day", col = "red", main = "Total Steps per Day (Frequency)", ylim = c(0,35))
```

#### 2.3 Calculation of the mean and median of the total steps per day

## 
```{r}
mean_median_and_other_values <- summary(total_steps_per_day_data$steps_per_day)
print(mean_median_and_other_values)
```

## What is the average daily activity pattern?
### Step 3: Identification of the average daily activity pattern

#### 3.1 Make a time series plot (i.e. type = "l" ) of the 5-minute interval 
#### (x-axis) and the average number of steps taken, averaged across all days (y-axis)

##### 3.1.1 Calculation of the average number of steps taken, averaged across all days for each 
##### 5-minute interval.
##### The value assigned to the NA elements (missing values) is zero in this implementation
##### This is accomplished through the na.rm = TRUE argument

```{r}
average_steps_per_interval_data <- data %>% 
        group_by(interval) %>% 
        summarise(steps_per_interval = mean(steps, na.rm = TRUE)) %>%
        ungroup
head(average_steps_per_interval_data)
tail(average_steps_per_interval_data)
str(average_steps_per_interval_data)
head(average_steps_per_interval_data$steps_per_interval)
```

##### 3.1.2 Creation of the plot of the 5-minute interval (x-axis) 
##### and the average number of steps taken, averaged across all days (y-axis)

```{r}
par(cex.lab = 0.8)
plot(average_steps_per_interval_data$interval, average_steps_per_interval_data$steps_per_interval, ylab = "Number of steps averaged across all days", xlab = "Interval", type = "l", pch = 3, lwd =1, main = "Number of steps averaged across all days \n for each 5 minute interval", col = "blue")
```

#### 3.2 Identification of the 5-minute interval, on average across all the days in the dataset, 
#### that contains the maximum number of steps

```{r}
interval_with_max_n_step_index <- which.max(average_steps_per_interval_data$steps_per_interval)
interval_with_max_n_step_index
```

#### Interval with the maximum number of steps

```{r}
interval_with_max_n_steps <- average_steps_per_interval_data$interval[interval_with_max_n_step_index]
interval_with_max_n_steps
```

#### check value and surrounding values

```{r}
average_steps_per_interval_data[100:106,]
```

## Imputing missing values
### Step 4: Inputting the missing  values

#### 4.1 Calculation and report of the total number of missing values in the dataset 
#### (i.e., the total number of rows with NAs)

```{r}
total_na_values <- sum(is.na(data$steps))

print(total_na_values)
```

#### 4.2 Filling of all the missing values in the dataset. 
#### It is possible to use the mean/median for that day, or the mean for that 
#### 5-minute interval, etc. to accomplish this objective.
#### In this implementation, the mean number of steps for each interval that was 
#### calculated without the NA values will be used to fill in the missing values
#### for the interval.


#### 4.3 Creation of a new dataset that is equal to the original dataset, but with the missing data #### filled in
##### 4.3.1 Add to the data a column with the mean for a specific interval

```{r}
data_with_average_steps_per_interval_data <- data %>% 
        group_by(interval) %>% 
        mutate(average_steps_per_interval = mean(steps, na.rm = TRUE)) %>%
        ungroup
head(data_with_average_steps_per_interval_data)
tail(data_with_average_steps_per_interval_data)
str(data_with_average_steps_per_interval_data)
```

##### 4.3.2 Selection of the rows with a steps value == NA

```{r}
na_values_selector <- is.na(data_with_average_steps_per_interval_data$steps)
head(na_values_selector)
```

##### 4.3.3 For these rows set the values in the steps column so that they are equal 
##### to the values in the column average_steps_per_interval

```{r}
data_with_average_steps_per_interval_data$steps[na_values_selector] <- data_with_average_steps_per_interval_data$average_steps_per_interval[na_values_selector]

head(data_with_average_steps_per_interval_data$steps[na_values_selector])
```

##### 4.3.4 Verification to check if there are any NA values left 

```{r}
total_na_values_after_change <- sum(is.na(data_with_average_steps_per_interval_data$steps))

print(total_na_values_after_change)
```

##### 4.3.5 Assignment of a new name to data

```{r}
data_without_na <- data_with_average_steps_per_interval_data
head(data_without_na)
```

#### 4.4 Creation of a histogram of the total number of steps taken each day 
#### and calculation and report of the mean and median total number of steps 
#### taken per day

##### 4.4.1 Calculation of sum of the steps per day for data without NA values
##### (i.e., for data in which the NA values have been changed into calcualted
##### values)

```{r}

total_steps_per_day_data_without_na <- data_without_na %>% 
        group_by(date) %>% 
        summarise(steps_per_day = sum(steps, na.rm = TRUE)) %>%
        ungroup
head(total_steps_per_day_data_without_na)
str(total_steps_per_day_data_without_na)
head(total_steps_per_day_data_without_na$steps_per_day)
```

##### 4.4.2 Creation of a histogram of the number of steps

```{r}
hist(total_steps_per_day_data_without_na$steps_per_day, breaks = 8, xlab = "Total Steps per Day", col = "red", main = "Total Steps per Day")
```

##### 4.4.3 Calculation of the mean and median of the total steps per day

## 
```{r}
mean_median_and_other_values_no_na <- summary(total_steps_per_day_data_without_na$steps_per_day)
print(mean_median_and_other_values_no_na)
```


###### 4.4.3.1 The values are more centered around the median after the elimination of the NA values,
###### and the mean is at a higher value. Both the median and the mean have increased.


##### 4.4.4 Analysis of how these values differ from the estimates in the first part of the  
##### assignment.
##### Analysis of the impact of imputing missing data on the estimates of the total daily number of ##### steps.

###### 4.4.4.1 Creation of a dataset with the data with NA values and without NA values

###### 4.4.4.1.1 Renaming of the column for the data without NA

```{r}
total_steps_per_day_data_without_na <- transmute(total_steps_per_day_data_without_na, date = date, steps_per_day_without_na = steps_per_day)
total_steps_per_day_data_without_na
```

###### 4.4.4.1.2 Creation of a dataset for the comparison

```{r}
dim(total_steps_per_day_data)
dim(total_steps_per_day_data_without_na)
comparison_total_steps_per_day_data <- left_join(total_steps_per_day_data, total_steps_per_day_data_without_na, by ="date")
head(comparison_total_steps_per_day_data)
```

###### 4.4.4.1.2 Modification of the dataset so there is a factor for the presence of NA

```{r}
comparison_total_steps_per_day_with_factor_data <- gather(comparison_total_steps_per_day_data, na_presence, steps, -date)
head(comparison_total_steps_per_day_with_factor_data)
str(comparison_total_steps_per_day_with_factor_data)
```


###### 4.4.4.1.3 Creation of a plot to show the difference between data with NA values and data without ###### NA values
##

```{r}
g <- ggplot(comparison_total_steps_per_day_with_factor_data, aes(steps, color = na_presence))
q <- g + geom_histogram(binwidth = 3000) + facet_grid(na_presence ~ .) + labs(title = "Total number of steps \ntaken each day (count)")
print(q)
```

###### 4.4.4.1.4 Comparison of two boxplots with NA values and without NA values

```{r}
par(mfrow = c(1,1))
boxplot(steps ~ na_presence, data = comparison_total_steps_per_day_with_factor_data, main="Comparison of Steps per Day between data that \nincludes NA values and data that does not")
```
###### 4.4.4.1.5 Comparison of median and mean values between data with NA values and data without NA ###### values

###### Results for the data with NA values
##

```{r}
mean_median_and_other_values <- summary(total_steps_per_day_data$steps_per_day)
print(mean_median_and_other_values)
```

###### Results for the data without NA values
## 

```{r}
mean_median_and_other_values_no_na <- summary(total_steps_per_day_data_without_na$steps_per_day)
print(mean_median_and_other_values_no_na)
```

## Are there differences in activity patterns between weekdays and weekends?
### Step 5: Find if there are differences in activity patterns between weekdays and weekends

###### For this step, the data without NA will be used.

```{r}
head(data_without_na)
str(data_without_na)
```

#### 5.1 Creation of a new factor variable in the dataset with two levels, “weekday” and “weekend” 
#### indicating whether a given date is a weekday or weekend day
  

##### 5.1.1 Modification of a date to change it into a date type

```{r}
data_without_na$date <- ymd(data_without_na$date)
str(data_without_na)
```

##### 5.1.2 Addition of a column with weekdays

```{r}

data_without_na_with_weekdays <- mutate(data_without_na, day_of_the_week = weekdays(date))
head(data_without_na_with_weekdays)
str(data_without_na_with_weekdays)
```

##### 5.1.3 Creation of another column with a factor

###### 5.1.3.1 Selection of the weekdays

```{r}

weekday_or_weekend <- ifelse(data_without_na_with_weekdays$day_of_the_week %in% c("Saturday", "Sunday"), "weekend", "weekday")
head(weekday_or_weekend)
length(weekday_or_weekend)
str(weekday_or_weekend)
```

###### 5.1.3.2 Addition of a column with weekday_or_weekend

```{r}
data_without_na_with_weekdays_and_weekday_weekend_factor <- mutate(data_without_na_with_weekdays, weekday_or_weekend = weekday_or_weekend)
head(data_without_na_with_weekdays_and_weekday_weekend_factor)
tail(data_without_na_with_weekdays_and_weekday_weekend_factor)
str(data_without_na_with_weekdays_and_weekday_weekend_factor)
```

###### 5.1.3.3 Modification of weekday_or_weekend to change it into a factor

```{r}
data_without_na_with_weekdays_and_weekday_weekend_factor$weekday_or_weekend <- as.factor(data_without_na_with_weekdays_and_weekday_weekend_factor$weekday_or_weekend)

str(data_without_na_with_weekdays_and_weekday_weekend_factor)
head(data_without_na_with_weekdays_and_weekday_weekend_factor)
tail(data_without_na_with_weekdays_and_weekday_weekend_factor)
```

#### 5.2 Creation of a panel plot containing a time series plot (i.e. type = "l" ) of the 5-minute 
#### interval (x-axis) and the average number of steps taken, averaged across all weekday days or 
#### weekend days (y-axis)

#####  5.2.1 Calculation of the average number of steps taken, averaged across all days for each 5-minute interval

```{r}

average_steps_per_interval_data <- data_without_na_with_weekdays_and_weekday_weekend_factor %>% 
        group_by(interval, weekday_or_weekend) %>% 
        summarise(steps_per_interval = mean(steps, na.rm = TRUE)) %>%
        ungroup
head(average_steps_per_interval_data)
tail(average_steps_per_interval_data)
str(average_steps_per_interval_data)
head(average_steps_per_interval_data$steps_per_interval)
tail(average_steps_per_interval_data$steps_per_interval)
str(average_steps_per_interval_data$steps_per_interval)
```

##### 5.2.2 Creation of the plot of the 5-minute interval (x-axis) 
##### and the average number of steps taken, averaged across all days (y-axis)
## 

```{r}
xyplot(steps_per_interval ~ interval | weekday_or_weekend,  data = average_steps_per_interval_data, layout = c(1,2), type = "l", main = "Average Number of Steps Averaged Across All Days", xlab = "Interval", ylab = "Steps per Interval")
```
